---
title: "Cleaning SBA Loan Data Set"
author: "Yash Tambawala"
date: "March 9, 2017"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, include=FALSE}
# Starting the libraries
library(dplyr)
library(ROCR)
library(data.table)
```


# Input

## Reading the files 

We read the following data sets:
1. Unemployment Data Set
2. FED Interest Rate data per year. 
Source: https://fred.stlouisfed.org/series/DFF/downloaddata
3. Total economic cost due to Tornados per year per state.
Source: http://www.spc.noaa.gov/wcm/test.html (manipulated using R and excel)
4. Small Business Lending Stats per state for 2011
Source:
https://www.sba.gov/advocacy/firm-size-data  https://www.sba.gov/content/small-business-lending-united-states-2010-2011


```{r}
set.seed(1)
df <- read.csv("SBA_loan_data_new.csv")

# Macroeconomic data sets 
unemployment_data <- read.csv("emp-unemployment.csv")
interest_rates_data <- read.csv("FED_IR.csv")
Tornado_damage_economic <- read.csv("Tornado FL.csv")
GDPSectorData <- read.csv("GDP_Industry.csv")
TED <- read.csv("TEDRATE.csv")
SNP500 <- read.csv("SNP500.csv")
```

## Creating the GDP data column

```{r}
set.seed(1)
GDP <- c(rep(0,length(df$ApprovalFiscalYear)))
```

# Data Cleaning operations

We do the following things:

1. Average HPI Index by State for 1990-2016 and adding PR,GU,DC and VI
2. Filtering out Loans canceled or exempt
3. Unemployment data modification

```{r}
set.seed(1)
# Average HPI Index by State for 1990-2016
# For GU, VI, DC, PR, we used the national averge HPI
hpi_state <- read.csv("hpi_state.csv", header=TRUE, stringsAsFactors=F)
hpi_state$ProjectState = as.factor(hpi_state$ProjectState)
df <- left_join(df, hpi_state, by=c("ProjectState", "ApprovalFiscalYear"))
df$mn_hpi = as.numeric(df$mn_hpi)

hpi_year_mean <- hpi_state %>% 
  group_by(ApprovalFiscalYear) %>%
  summarise(mn_hpi = mean(mn_hpi))
hpi_year_mean$ProjectState <- "PR"
hpi_year_mean <- hpi_year_mean[,c(3,1,2)]
hpi_state <- rbind(hpi_state,hpi_year_mean)
hpi_year_mean$ProjectState <- "GU"
hpi_state <- rbind(hpi_state,hpi_year_mean)
hpi_year_mean$ProjectState <- "DC"
hpi_state <- rbind(hpi_state,hpi_year_mean)
hpi_year_mean$ProjectState <- "VI"
hpi_state <- rbind(hpi_state,hpi_year_mean)

temp <- transpose(as.data.frame(rep(0,29))) 
colnames(temp) <- colnames(unemployment_data)
unemployment_data <- rbind(unemployment_data,temp)
unemployment_data$X <- as.character(unemployment_data$X)
unemployment_data[52,3] <- "PR"
unemployment_data <- rbind(unemployment_data,temp)
unemployment_data[53,3] <- "GU"
unemployment_data <- rbind(unemployment_data,temp)
unemployment_data[54,3] <- "DC"
unemployment_data <- rbind(unemployment_data,temp)
unemployment_data[55,3] <- "VI"

for(j in 52:55){
 for(i in 4:29){
  unemployment_data[j,i] <- mean(unemployment_data[2:51,i])
}
}
```

Filtering out Loans which are canceled or exempt
```{r}
df <- filter(df, LoanStatus != "CANCLD")
df <- filter(df, LoanStatus != "EXEMPT")
df <- filter(df, DeliveryMethod != "504REFI")
```

We now remove extra columns

1. Removing the first 3 columns of the data set  as they contain the common program info and a unique identifying name for every business
2. Removing the column BorrZip Code
3. Removing CDC street, city, zip
4. Adding 4 columns - isDefault, NotSameState, ThirdPartyApproved and dayselapsed
5. Conversion to factor
6. Removing the Third Party Lender Name, City, State
7. Removing subpgmdesc
8. Removing NAICS description
9. Removing Project County and State
10. Adding 2 digits NAICS code

```{r}
set.seed(1)
df <- subset(df,select = -c(Program,BorrName,BorrStreet,
                     BorrZip,CDC_Street,CDC_City,CDC_Zip))

# adding isDefault

df$isDefault <- ifelse(df$LoanStatus=="CHGOFF",1,0)
for (i in 1:nrow(df)) {
if(df$LoanStatus[i]=="PIF"){df$isDefault[i]= 0}} 
df$isDefault <- as.numeric(df$isDefault)

# factor conversion
df$NotSameState <- factor(df$NotSameState)
df$ThirdPartyApproved <- factor(df$ThirdPartyApproved)

# adding dayselaped
df <- mutate(df,dayselapsed = as.numeric(difftime(strptime(ChargeOffDate, format = "%m/%d/%Y"),
                                                  strptime(ApprovalDate, format = "%m/%d/%Y"))))

df$dayselapsed[is.na(df$dayselapsed)]=round(df$TermInMonths*30.4)
df$dayselapsed[is.na(df$dayselapsed)] <- 7300

# adding FinalYear
df$FinalYear=df$ApprovalFiscalYear+round((df$dayselapsed)/365)

# Removing extra columns
df <- subset(df,select = -c(ThirdPartyLender_Name,ThirdPartyLender_City,ThirdPartyLender_State,
                     subpgmdesc,NaicsDescription,ProjectCounty,ProjectState))

# adding Naics2digits
df <- mutate(df,Naics2digits = substr(NaicsCode,1,2))

```

We perform the following operations:


1. Getting Interest Rates table

```{r}
# Getting Interest Rates
set.seed(1)
interest_rates=matrix(0,nrow=length(df$ApprovalFiscalYear),ncol=length(1990:2014))
for (i in 1:length(1990:2014)) {
  interest_rates[,i]=interest_rates_data[match(1989+i,interest_rates_data$Year),2]
}
interest_rates = interest_rates[1,]

```

2. Discretize the data into 1-year period

```{r}
  # Initializing Matrix for Adding columns
  set.seed(1)
  matrix_test=matrix(0,nrow=0,ncol=23)
  finalMatrix=matrix(0,nrow=0,ncol=23)
  tempRow = matrix(0,nrow=1,ncol=23)
  
  end_index=0
  
  x=c("loanNum","start","stop","interestRate","gdpIndustry","unemploymentRate","hpiState","tedSpread","sandp500",
      "BorrState","CDC_State","ThirdPartyDollars","GrossApproval","ApprovalDate","ApprovalFiscalYear","DeliveryMethod",
      "TermInMonths","BusinessType","NotSameState","Naics2digits","Defaultyn","isDefault","endIndicator")
  
  colnames(matrix_test)=x
  
  df_const <- as.matrix(subset(df,select = c(BorrState,CDC_State,ThirdPartyDollars,GrossApproval,
                                             ApprovalDate,ApprovalFiscalYear,DeliveryMethod,
                                             TermInMonths,BusinessType,NotSameState,Naics2digits,isDefault)))
  
  for (i in 1:nrow(df)) {
    start=df$ApprovalFiscalYear[i]
    stop=df$ApprovalFiscalYear[i]+round(df$dayselapsed[i]/365)
    periods=length(start:stop)     
    n=0
    unemployment_start_index=match(start,unemployment_data[1,])
    unemployment_state_index=match(df$BorrState[i],unemployment_data[,3])
    if(is.na(df$Naics2digits[i])|df$Naics2digits[i]==99){
      GDP_NAICS_index=2
    }
    else{
      GDP_NAICS_index=match(df$Naics2digits[i],GDPSectorData[,2])
    }
    GDP_start_index=match(start,GDPSectorData[1,])
    HPI_start_index=match(interaction(df$BorrState[i],start),
                          interaction(hpi_state$ProjectState, hpi_state$ApprovalFiscalYear))
    
    if(stop<=2014){
      for (j in 1:periods) { 
        tempRow = matrix(0,nrow=1,ncol=23)
        tempRow[1,1]=i
        tempRow[1,2]=start+n
        tempRow[1,3]=start+n+1
        tempRow[1,4] = interest_rates[start-1990+j]
        tempRow[1,5]=GDPSectorData[GDP_NAICS_index,GDP_start_index+n]
        tempRow[1,6]=unemployment_data[unemployment_state_index,unemployment_start_index+n]
        tempRow[1,7]=hpi_state[HPI_start_index+n,3]
        tempRow[1,8]=TED[start-1990+j,2]
        tempRow[1,9]=SNP500[start-1990+j,1]
        tempRow[1,as.numeric(10:21)] = df_const[i,]
        # endIndicator Initialisation
        if(df$isDefault[i]==1){
          if((start+n)==stop){tempRow[1,22]=1
          tempRow[1,23]=1}
        }
        
        if(start+n==stop) {tempRow[1,23]=1}
        
        n=n+1 
        matrix_test <- rbind(matrix_test,tempRow)
      }
    
      
    }
    else{
      for (j in 1:(2014-start)) {
        tempRow = matrix(0,nrow=1,ncol=23)
        tempRow[1,1]=i
        tempRow[1,2]=start+n
        tempRow[1,3]=start+n+1
        tempRow[1,4]=interest_rates[start-1990+j]
        tempRow[1,5]=GDPSectorData[GDP_NAICS_index,GDP_start_index+n]
        tempRow[1,6]=unemployment_data[unemployment_state_index,unemployment_start_index+n]
        tempRow[1,7]=hpi_state[HPI_start_index+n,3]
        tempRow[1,8]=TED[start-1990+j,2]
        tempRow[1,9]=SNP500[start-1990+j,1]
        tempRow[1,10:21] = df_const[i,]
        # endIndicator Initialisation
        if(df$isDefault[i]==1){
          if((start+n)==stop){tempRow[1,22]=1
          tempRow[1,23]=1}
        }
        
        if(unemployment_start_index+n==stop) {tempRow[1,23]=1}
        
        if(j==2014-start){tempRow[1,23]=1}
        
        n=n+1 
        matrix_test <- rbind(matrix_test,tempRow)
      }
    }
    if(nrow(matrix_test)>=1000){
      finalMatrix <- rbind(finalMatrix,matrix_test)
      matrix_test=matrix(0,nrow=0,ncol=23)
    }
  }
  

finalMatrix <- rbind(finalMatrix,matrix_test)
finalMatrix=as.data.frame(finalMatrix)

# Convert ApprovalDate
finalMatrix <- mutate(finalMatrix, ApprovalDate = as.numeric(difftime(strptime(ApprovalDate, format ="%m/%d/%Y"),strptime("1/1/1990", format = "%m/%d/%Y"))))

# Store these data as numeric
finalMatrix$start <- as.numeric(as.character(finalMatrix$start))
finalMatrix$stop <- as.numeric(as.character(finalMatrix$stop))
finalMatrix$isDefault <- as.numeric(as.character(finalMatrix$isDefault))
finalMatrix$GrossApproval <- as.numeric(as.character(finalMatrix$GrossApproval))
finalMatrix$unemploymentRate <- as.numeric(as.character(finalMatrix$unemploymentRate))
finalMatrix$interestRate <- as.numeric(as.character(finalMatrix$interestRate))
finalMatrix$hpiState <- as.numeric(as.character(finalMatrix$hpiState))
finalMatrix$TermInMonths <- as.numeric(as.character(finalMatrix$TermInMonths))
finalMatrix$tedSpread <-as.numeric(as.character(finalMatrix$tedSpread))
finalMatrix$sandp500 <- as.numeric(as.character(finalMatrix$sandp500))
finalMatrix$loanNum <- as.numeric(as.character(finalMatrix$loanNum))
finalMatrix$ThirdPartyDollars <- as.numeric(as.character(finalMatrix$ThirdPartyDollars))

# Replace NAs in ThirdPArtyDollars with 0
finalMatrix$ThirdPartyDollars[is.na(finalMatrix$ThirdPartyDollars)]<- 0

# Replace NAs in Naics2digits with 0
finalMatrix$Naics2digits <- as.numeric(as.character(finalMatrix$Naics2digits))
finalMatrix$Naics2digits[is.na(finalMatrix$Naics2digits)]<- 0
finalMatrix$Naics2digits <- as.factor(finalMatrix$Naics2digits)

# Export file
write.csv(finalMatrix, file="big_data_Matrix.csv", col.names=FALSE)

```