---
title: "logistic regression"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(dplyr)
library(ROCR)
library(glmnet)
library(leaps)
library(randomForest)
```

#Data Input

```{r}
set.seed(1)
training_set = read.csv("training_set.csv")
val_set = read.csv("validation_set.csv")
test500_full = read.csv("test500_full (1).csv")
test500_full$Naics2digits = as.factor(test500_full$Naics2digits)
training_set$Naics2digits = as.factor(training_set$Naics2digits)
val_set$Naics2digits = as.factor(val_set$Naics2digits)
```

#Filtering first year data

``` {r}
# filter first year
training_set_1yr = filter(training_set, ApprovalFiscalYear == start)
val_set_1yr = filter(val_set, ApprovalFiscalYear == start)

# remove isDefault and rename Defaultyn as isDefault
# because we only care about whether the loan defaulted, the time of default is not what we need to know
drop = c("isDefault")
training_set_1yr = training_set_1yr[, !(names(training_set_1yr) %in% drop)]
colnames(training_set_1yr)[23] = "isDefault"
val_set_1yr = val_set_1yr[, !(names(val_set_1yr) %in% drop)]
colnames(val_set_1yr)[23] = "isDefault"
# Change Naics2digits from numeric to factor
training_set_1yr$Naics2digits = as.factor(training_set_1yr$Naics2digits)
val_set_1yr$Naics2digits = as.factor(val_set_1yr$Naics2digits)
```

#Lasso for baseline

```{r}
lasso_training_1yr = training_set_1yr
x.lasso = model.matrix(lasso_training_1yr$isDefault~GrossApproval+hpiState+unemploymentRate+interestRate+sandp500+NotSameState+BusinessType+DeliveryMethod+TermInMonths+ApprovalFiscalYear+BorrState+Naics2digits+gdpIndustry+tedSpread+ThirdPartyDollars, data = lasso_training_1yr)[,-1]
lasso.mod = glmnet(x.lasso, lasso_training_1yr$isDefault, alpha = 1)
cv.out = cv.glmnet(x.lasso, lasso_training_1yr$isDefault, alpha = 1)
plot(cv.out)
cv.out
bestlam=cv.out$lambda.min
predict(lasso.mod, s = cv.out$lambda.min, type = "coefficients")
predict(lasso.mod, s = cv.out$lambda.1se, type = "coefficients")

#no macro economic variables
x.lasso_no_time = model.matrix(lasso_training_1yr$isDefault~GrossApproval+NotSameState+BusinessType+DeliveryMethod+TermInMonths+ApprovalFiscalYear+BorrState+Naics2digits+ThirdPartyDollars, data = lasso_training_1yr)[,-1]
lasso.mod_no_time = glmnet(x.lasso_no_time, lasso_training_1yr$isDefault, alpha = 1)
cv.out_no_time = cv.glmnet(x.lasso_no_time, lasso_training_1yr$isDefault, alpha = 1)
plot(cv.out_no_time)
cv.out_no_time
bestlam=cv.out$lambda.min
predict(lasso.mod_no_time, s = cv.out_no_time$lambda.min, type = "coefficients")
predict(lasso.mod_no_time, s = cv.out_no_time$lambda.1se, type = "coefficients")
```

#Variable importance for baseline

```{r}
var_imp_1yr = training_set_1yr
var_imp_1yr = filter(var_imp_1yr, CDC_State != "GU")
var_imp_1yr = filter(var_imp_1yr, CDC_State != "PR")
var_imp_1yr = filter(var_imp_1yr, CDC_State != "VI")
var_imp_1yr = filter(var_imp_1yr, BorrState != "GU")
var_imp_1yr = filter(var_imp_1yr, BorrState != "PR")
var_imp_1yr = filter(var_imp_1yr, BorrState != "VI")
var_imp_1yr = droplevels(var_imp_1yr)
var_imp_1yr$isDefault = as.factor(var_imp_1yr$isDefault)
var_imp_1yr$Naics2digits = as.factor(var_imp_1yr$Naics2digits)
rf.test_1yr =randomForest(isDefault~GrossApproval+hpiState+unemploymentRate+interestRate+sandp500+NotSameState+BusinessType+DeliveryMethod+TermInMonths+ApprovalFiscalYear+BorrState+Naics2digits+gdpIndustry+tedSpread+ThirdPartyDollars,data=var_imp_1yr,mtry = 4, importance = TRUE)
importance(rf.test_1yr)
varImpPlot(rf.test_1yr,type=2)

rf.test_no_time =randomForest(isDefault~GrossApproval+NotSameState+BusinessType+DeliveryMethod+TermInMonths+ApprovalFiscalYear+BorrState+Naics2digits+ThirdPartyDollars,data=var_imp_1yr,mtry = 3, importance = TRUE)
importance(rf.test_no_time)
varImpPlot(rf.test_no_time,type=2)
```

#Baseline with macro economic variables at the time loan initiated

```{r}

log_reg_1yr = glm(isDefault ~ log(GrossApproval)+hpiState+unemploymentRate+interestRate+sandp500+NotSameState+BusinessType+DeliveryMethod+TermInMonths+ApprovalFiscalYear+BorrState+Naics2digits+gdpIndustry+tedSpread+ThirdPartyDollars,family="binomial",data=training_set_1yr)
# val_set
pred_log_reg_1yr=predict(log_reg_1yr,val_set_1yr,type="response")

pred1yr = prediction(pred_log_reg_1yr, val_set_1yr$isDefault)
perf1yr = performance(pred1yr, "tpr", "fpr")
plot(perf1yr)
auc1yr <- performance(pred1yr,measure = "auc")
auc1yr <- auc1yr@y.values[[1]]
auc1yr
```

#Baseline without macro economic variables at the time loan initiated

```{r}
df.lm = glm(isDefault~log(GrossApproval)+BorrState+DeliveryMethod+ApprovalFiscalYear+BusinessType+NotSameState+ThirdPartyDollars+Naics2digits, family="binomial",data=training_set_1yr)

# val_set
df.pred = predict(df.lm, val_set_1yr, type = "response")
# ROC Curve
df.pred.auc = prediction(df.pred, val_set_1yr$isDefault)
perf.df = performance(df.pred.auc, "tpr", "fpr")
plot(perf.df)
auc.df <- performance(df.pred.auc,measure = "auc")
auc.df <- auc.df@y.values[[1]]
auc.df
```

#Lasso for logistic Regression

```{r Logistic Regression}
#Lasso
lasso_training = training_set
training_set$Naics2digits = as.factor(training_set$Naics2digits)
x.lasso = model.matrix(training_set$isDefault~GrossApproval+hpiState+unemploymentRate+interestRate+sandp500+NotSameState+BusinessType+DeliveryMethod+TermInMonths+ApprovalFiscalYear+BorrState+CDC_State+Naics2digits+gdpIndustry, data = training_set)[,-1]

lasso_training = filter(lasso_training, !is.na(unemploymentRate))
lasso.mod = glmnet(x.lasso, lasso_training$isDefault, alpha = 1)
cv.out = cv.glmnet(x.lasso, lasso_training$isDefault, alpha = 1)
plot(cv.out)
cv.out
predict(lasso.mod, s = cv.out$lambda.min, type = "coefficients")
predict(lasso.mod, s = cv.out$lambda.1se, type = "coefficients")
```

#vaiable importance for logistic regression model

```{r}
var_imp = training_set
var_imp = filter(var_imp, CDC_State != "GU")
var_imp = filter(var_imp, CDC_State != "PR")
var_imp = filter(var_imp, CDC_State != "VI")
var_imp = filter(var_imp, BorrState != "GU")
var_imp = filter(var_imp, BorrState != "PR")
var_imp = filter(var_imp, BorrState != "VI")
var_imp = droplevels(var_imp)
var_imp$isDefault = as.factor(var_imp$isDefault)
var_imp$Naics2digits = as.factor(var_imp$Naics2digits)
rf.test =randomForest(isDefault~GrossApproval+hpiState+unemploymentRate+interestRate+sandp500+NotSameState+BusinessType+DeliveryMethod+TermInMonths+ApprovalFiscalYear+BorrState+Naics2digits+gdpIndustry+tedSpread+ThirdPartyDollars,data=var_imp,mtry = 4, importance = TRUE)
importance(rf.test)
varImpPlot(rf.test,type=2)
```

#Model Construction

```{r}
log_reg=glm(isDefault ~ interestRate+ApprovalFiscalYear+gdpIndustry+
              log(GrossApproval)+hpiState+unemploymentRate+
              sandp500+NotSameState+Naics2digits+BorrState+tedSpread+ThirdPartyDollars
            ,family="binomial",data=training_set)
summary(log_reg)

# prediction of val_set
pred_log_reg=predict(log_reg,val_set,type="response")

# find start and stop for each loan
loan_indices <- as.data.frame(unique(val_set$loanNum))
loan_indices$start <- 1
loan_indices$stop <- 1
colnames(loan_indices) <- c("loan_no","start","stop")
l = 1

for(i in 1:nrow(val_set)){
  if(val_set$endIndicator[i]==1){
    loan_indices$stop[l]=i
    l=l+1
    if(l<=nrow(loan_indices)){
    loan_indices$start[l]=i+1
    }
  }
}

# survival probability of each loan in each year
# probability p = a loan survived from the start to the year X
# 1 - p = probability that a loan defaulted at some point between start and X
pred_test <- rep(0,length(pred_log_reg))
counter = 1

for(i in 1:nrow(loan_indices))
{
  a = loan_indices$start[i]
  b = loan_indices$stop[i]
  for(j in a:b){
    if(j==a){
      pred_test[a]=pred_log_reg[a]
    }
    else{
      for(k in a:j){
        counter = counter*(1-pred_log_reg[k])
      }
       pred_test[j]=1-counter
    }
    counter = 1
  }
}

#ROC Curve
pred1 = prediction(pred_test, val_set$isDefault)
perf1 = performance(pred1, "tpr", "fpr")
plot(perf1, col = "black")
auc1 <- performance(pred1,measure = "auc")
auc1 <- auc1@y.values[[1]]
auc1

plot(perf1yr, add = T, col = "red")
plot(perf.df, add = T, col = "blue")
legend("bottomright", legend = c("discretized data", "fixed 1-year macro economic variables", "no macro economic variables"), col = c("black", "red", "blue"), lty = 1, cex = 0.8)
```

#Logistic Regression on test 500 dataset

```{r}
# prediction of val_set
pred_log_reg=predict(log_reg,test500_full,type="response")

# find start and stop for each loan
loan_indices <- as.data.frame(unique(val_set$loanNum))
loan_indices$start <- 1
loan_indices$stop <- 1
colnames(loan_indices) <- c("loan_no","start","stop")
l = 1

for(i in 1:nrow(val_set)){
  if(val_set$endIndicator[i]==1){
    loan_indices$stop[l]=i
    l=l+1
    if(l<=nrow(loan_indices)){
    loan_indices$start[l]=i+1
    }
  }
}

# survival probability of each loan in each year
# probability p = a loan survived from the start to the year X
# 1 - p = probability that a loan defaulted at some point between start and X

pred_test <- rep(0,length(pred_log_reg))
counter = 1

for(i in 1:nrow(loan_indices))
{
  a = loan_indices$start[i]
  b = loan_indices$stop[i]
  for(j in a:b){
    if(j==a){
      pred_test[a]=pred_log_reg[a]
    }
    else{
      for(k in a:j){
        counter = counter*(1-pred_log_reg[k])
      }
       pred_test[j]=1-counter
    }
    counter = 1
  }
}

#ROC Curve
pred1 = prediction(pred_test, val_set$isDefault)
perf1 = performance(pred1, "tpr", "fpr")
plot(perf1)
auc1 <- performance(pred1,measure = "auc")
auc1 <- auc1@y.values[[1]]
auc1
```
